---
title: "RMeDPower2 Class Configuration Guide"
author: "RMeDPower2 Package"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)
```

# Overview

RMeDPower2 uses three S4 class objects to structure experimental designs and power analysis parameters. This guide provides decision criteria for configuring each class based on your experimental setup.

The three core classes are:

- **RMeDesign**: Defines your experimental structure and statistical model specifications
- **ProbabilityModel**: Specifies the statistical distribution of your response variable  
- **PowerParams**: Configures power analysis parameters

---

# RMeDesign Class

The `RMeDesign` class defines your experimental structure and statistical model specifications.

## Required Parameters

### `response_column`
**What it is:** Name of your outcome/dependent variable column  
**How to choose:** This should be the quantitative measure you're analyzing

```{r response_column_examples}
# Examples:
response_column = "cell_size"        # Cell morphology study
response_column = "expression_level" # Gene expression analysis
response_column = "reaction_time"    # Behavioral study
```

### `condition_column` 
**What it is:** Your main predictor/treatment variable  
**How to choose:** The primary experimental factor you want to test

```{r condition_column_examples}
# Examples:
condition_column = "treatment"       # Drug vs control
condition_column = "genotype"       # Wild-type vs mutant
condition_column = "time_point"     # Longitudinal measurements
```

### `condition_is_categorical`
**Decision criteria:**

- `TRUE`: Treatment groups, genotypes, categorical time points
- `FALSE`: Continuous measurements like dose, age, time as numeric

### `experimental_columns`
**What it is:** Hierarchical grouping variables (most general to most specific)  
**How to order:** Always list from highest to lowest level

```{r experimental_columns_examples}
# Examples:
experimental_columns = c("experiment", "plate", "well")     # Lab hierarchy
experimental_columns = c("batch", "animal", "sample")       # Animal study
experimental_columns = c("study_site", "participant")       # Clinical trial
```

## Optional Parameters

### `covariate`
**When to include:** You have a confounding variable to control for

```{r covariate_examples}
covariate = "age"           # Control for age effects
covariate = "batch_date"    # Control for batch effects
covariate = "baseline"      # Control for pre-treatment values
```

### `covariate_is_categorical`
**Decision criteria:**

- `TRUE`: Gender, batch labels, categorical age groups
- `FALSE`: Continuous age, baseline measurements, numeric scores

### `crossed_columns`
**When to use:** Some experimental units appear in multiple higher-level groups  
**Example:** Same cell line used across multiple experiments

```{r crossed_columns_example}
experimental_columns = c("experiment", "cell_line")
crossed_columns = "cell_line"  # Cell lines repeat across experiments
```

### `include_interaction`
**When to use:** You expect the covariate effect varies by condition

**Examples:**
- Drug effect varies by age
- Treatment response differs by genotype  
- Time effects depend on intervention group

### `random_slope_variable`
**When to use:** You expect the effect varies across experimental units

**Options:**
- `"condition_column"`: Treatment effects vary across subjects/units
- `"covariate"`: Covariate effects vary across experimental levels

---

# ProbabilityModel Class

Defines the statistical distribution of your response variable.

## `error_is_non_normal`
**Decision criteria:**

- `FALSE`: Continuous, approximately normal data (heights, measurements, expression levels)
- `TRUE`: Count data, proportions, or clearly non-normal distributions

## `family_p` (when `error_is_non_normal = TRUE`)

### `"poisson"`
- **Use for:** Count data (number of cells, events, mutations)
- **Characteristics:** Non-negative integers, variance ≈ mean
- **Example:** Number of colonies per plate

### `"binomial"` 
- **Use for:** Proportions or binary outcomes
- **Requires:** `total_column` specifying denominator
- **Example:** Number of positive cells out of total counted

### `"negative_binomial"`
- **Use for:** Overdispersed count data 
- **When:** Count data where variance > mean
- **Example:** RNA-seq read counts

### `"Gamma"`
- **Use for:** Positive continuous data with right skew
- **Example:** Reaction times, concentrations

---

# PowerParams Class

Configures power analysis parameters.

## Core Parameters

### `target_columns`
**What it is:** Which experimental factors to vary in power analysis  
**How to choose:** The bottleneck factor(s) in your design

```{r target_columns_examples}
target_columns = "experiment"     # Want more experimental replicates
target_columns = "animal"         # Need more animals per group
target_columns = c("experiment", "animal")  # Optimize both levels
```

### `levels`
**Decision criteria for each target column:**

- `1`: Increase number of groups/levels (more experiments, more animals)
- `0`: Increase sample size within existing groups (more samples per animal)

### `power_curve`
- `1`: Generate power curves showing power vs. sample size
- `0`: Calculate power for specific sample size

## Sample Size Parameters

### `max_size`
**Default:** Current size × 5  
**When to specify:** You have budget/feasibility constraints

```{r max_size_example}
max_size = c(10, 50)  # Max 10 experiments, 50 animals per group
```

### `nsimn`
**Default:** 1000 simulations  
**When to change:**

- Increase for more precision (slower)
- Decrease for faster exploratory analysis

## Effect Size Specification

### `effect_size`
**When to specify:** You know the biologically meaningful effect size  
**How to determine:**

- From pilot studies
- From literature
- From biological significance criteria

### `ICC` (Intra-Class Correlation)
**When to use:** You have prior knowledge of experimental variability  
**How to specify:** Vector matching `experimental_columns` order

---

# Complete Configuration Examples

## Example 1: Cell Culture Experiment

```{r cell_culture_example}
library(RMeDPower2)

# Design: Test drug effect on cell size across experiments and plates
design <- new("RMeDesign",
  response_column = "cell_area",
  condition_column = "treatment",           # "control" vs "drug"
  condition_is_categorical = TRUE,
  experimental_columns = c("experiment", "plate"),
  crossed_columns = NULL,                   # Plates are nested in experiments
  covariate = "passage_number",
  covariate_is_categorical = FALSE,         # Continuous variable
  include_interaction = FALSE
)

model <- new("ProbabilityModel",
  error_is_non_normal = FALSE              # Cell area is approximately normal
)

power_params <- new("PowerParams",
  target_columns = "experiment",           # Want to know how many experiments needed
  levels = 1,                             # Add more experiments (not more plates per experiment)
  power_curve = 1,                        # Generate power curve
  nsimn = 1000,
  alpha = 0.05
)

# Run power analysis
# result <- calculatePower(data = your_data, design = design, 
#                          model = model, power_param = power_params)
```

## Example 2: RNA-seq Count Data

```{r rnaseq_example}
# Design: Compare gene expression between genotypes with batch effects
design <- new("RMeDesign",
  response_column = "read_count",
  condition_column = "genotype",
  condition_is_categorical = TRUE,
  experimental_columns = c("batch", "animal"),
  total_column = "total_reads",            # For normalization
  covariate = NULL
)

model <- new("ProbabilityModel",
  error_is_non_normal = TRUE,
  family_p = "negative_binomial"           # Overdispersed count data
)

power_params <- new("PowerParams",
  target_columns = "animal",
  levels = 1,                             # Add more animals
  power_curve = 1,
  alpha = 0.05
)

# Run power analysis
# result <- calculatePower(data = your_data, design = design, 
#                          model = model, power_param = power_params)
```

## Example 3: Behavioral Study with Continuous Predictor

```{r behavioral_example}
# Design: Test dose-response relationship in reaction time study
design <- new("RMeDesign",
  response_column = "reaction_time",
  condition_column = "dose",               # Continuous dose levels
  condition_is_categorical = FALSE,
  experimental_columns = c("session", "subject"),
  covariate = "age",
  covariate_is_categorical = FALSE,
  include_interaction = TRUE,              # Dose effect may vary by age
  random_slope_variable = "condition_column" # Dose effects vary by subject
)

model <- new("ProbabilityModel",
  error_is_non_normal = TRUE,
  family_p = "Gamma"                      # Reaction times are right-skewed
)

power_params <- new("PowerParams",
  target_columns = "subject",
  levels = 1,
  power_curve = 1,
  effect_size = 0.5                       # Known meaningful effect size
)

# Run power analysis
# result <- calculatePower(data = your_data, design = design, 
#                          model = model, power_param = power_params)
```

## Example 4: Clinical Trial with Categorical Covariate

```{r clinical_example}
# Design: Multi-site clinical trial with gender as covariate
design <- new("RMeDesign",
  response_column = "symptom_score",
  condition_column = "treatment_group",
  condition_is_categorical = TRUE,
  experimental_columns = c("site", "participant"),
  covariate = "gender",
  covariate_is_categorical = TRUE,         # Categorical covariate
  include_interaction = TRUE,              # Treatment effect may differ by gender
  crossed_columns = NULL
)

model <- new("ProbabilityModel",
  error_is_non_normal = FALSE              # Symptom scores approximately normal
)

power_params <- new("PowerParams",
  target_columns = c("site", "participant"),
  levels = c(1, 1),                       # Optimize both sites and participants
  power_curve = 1,
  nsimn = 500,                            # Reduce for faster computation
  alpha = 0.05
)

# Run power analysis
# result <- calculatePower(data = your_data, design = design, 
#                          model = model, power_param = power_params)
```

---

# Decision Flowchart

## Step 1: Data Structure
1. **What's your outcome variable?** → `response_column`
2. **What's your main comparison?** → `condition_column`
3. **What are your grouping levels?** → `experimental_columns`

## Step 2: Statistical Considerations
1. **Is your outcome normally distributed?** → `error_is_non_normal`
2. **Do you have covariates to control for?** → `covariate`, `include_interaction`
3. **Do effects vary across groups?** → `random_slope_variable`
4. **Is your covariate categorical or continuous?** → `covariate_is_categorical`

## Step 3: Power Analysis Goals
1. **What's your limiting factor?** → `target_columns`
2. **Do you want more groups or bigger groups?** → `levels`
3. **What's your budget constraint?** → `max_size`

---

# Tips for Success

## Data Preparation
- Ensure column names match exactly what you specify in the design objects
- Check that categorical variables are properly coded
- Verify hierarchical structure is correctly represented

## Model Selection
- Use diagnostic plots to verify distributional assumptions
- Consider the `diagnoseDataModel()` function to check model appropriateness
- Start with simpler models before adding interactions

## Power Analysis Strategy
- Begin with smaller `nsimn` values for exploratory analysis
- Focus on one `target_column` at a time initially
- Use pilot data or literature to inform effect size estimates

## Troubleshooting
- If power analysis fails, check that your design matches your data structure
- Ensure sufficient data for the complexity of your model
- Verify that crossed vs. nested experimental factors are correctly specified

---

# Additional Resources

## Package Functions
- `calculatePower()`: Main function for power analysis
- `diagnoseDataModel()`: Check model assumptions and identify outliers
- `get_model_and_data()`: Fit models and examine results
- `check_normality()`: Assess normality of residuals

## Input Templates
The package includes JSON templates in the `input_templates/` directory:
- `design_template.json`: Template for RMeDesign parameters
- `power_param_template.json`: Template for PowerParams parameters  
- `stat_model_template.json`: Template for ProbabilityModel parameters

## Example Data
Use the built-in example datasets (`RMeDPower_data1` through `RMeDPower_data7`) to practice with different experimental designs and learn the package functionality.

```{r load_example_data, eval=TRUE}
# Load and examine example data
library(RMeDPower2)
data(RMeDPower_data1)
head(RMeDPower_data1)
str(RMeDPower_data1)
```

This framework ensures your class objects accurately represent your experimental design and analysis goals, leading to meaningful and reliable power analysis results.

---

# Interactive Configuration with Shiny App

## Overview

The RMeDPower2 package includes an interactive Shiny web application that guides users through the process of creating JSON configuration files for the three S4 classes. This app provides a user-friendly interface with data-driven suggestions and real-time validation.

## Getting Started with the Shiny App

### Installation and Setup

```{r shiny_setup, eval=FALSE}
# Navigate to the Shiny app directory
setwd("path/to/RMeDPower2/shiny_config_app")

# Install required packages
source("install_packages.R")

# Or install manually:
install.packages(c("shiny", "shinydashboard", "DT", "jsonlite"))
```

### Running the App

```{r run_shiny_app, eval=FALSE}
# Run the RMeDPower2 Configuration App
shiny::runApp("app.R")
```

## App Workflow

### Step 1: Data Upload

1. **Upload your data file** (CSV or RDS format)
2. **Review the enhanced data preview** showing:
   - Each column on a separate line with data type
   - Sample values from the first 5 rows
   - Dataset dimensions summary

```
Data Preview (first 5 rows):
==================================================

response             (numeric): 1.234, 2.567, 3.891, 4.123, 5.456
condition            (character): A, B, A, B, A
experiment           (numeric): 1, 1, 2, 2, 3
plate                (numeric): 1, 2, 1, 2, 1
batch                (character): batch1, batch1, batch2, batch2, batch3

Dataset: 5 rows × 6 columns
```

### Step 2: RMeDesign Configuration

The app automatically populates column choices from your uploaded data:

#### Required Parameters
- **Response Column**: Select your outcome/dependent variable
- **Condition Column**: Choose your main predictor/treatment variable  
- **Condition Type**: Specify if categorical or continuous
- **Experimental Hierarchy**: Select grouping variables (highest to lowest level)

#### Optional Parameters
- **Covariate**: Add confounding variables to control for
- **Crossed Columns**: Specify variables that repeat across hierarchy levels
- **Total Column**: For binomial data, specify the denominator column
- **Random Slope Variable**: Allow effects to vary across experimental units

#### Smart Defaults
If parameters aren't specified, the app provides sensible defaults:
- First column → response_column
- Second column → condition_column  
- Remaining columns → experimental_columns
- Categorical conditions (most common case)
- No interactions initially

### Step 3: ProbabilityModel Configuration

#### Distribution Selection
- **Normal Distribution**: For continuous, approximately normal data
- **Non-Normal Distribution**: For counts, proportions, or skewed data

#### Family Selection (for non-normal data)
When non-normal is selected, a dropdown appears with options:

- **Poisson**: Count data where variance ≈ mean
  - Example: Number of colonies, events, mutations
- **Negative Binomial**: Overdispersed count data (variance > mean)
  - Example: RNA-seq read counts
- **Binomial**: Proportions or binary outcomes
  - Example: Number of positive cells out of total counted
  - **Requires**: total_column in RMeDesign
- **Gamma**: Positive continuous data with right skew
  - Example: Reaction times, concentrations

#### Automatic Parameter Assignment
The app intelligently handles the `family_p` parameter:
- Normal distribution → `family_p = NULL`
- Non-normal selected but no family chosen → `family_p = "poisson"` (default)
- Specific family selected → `family_p = selected_family`

### Step 4: PowerParams Configuration

#### Core Parameters
- **Target Columns**: Which experimental factors to vary in power analysis
- **Levels**: For each target column (1 = more groups, 0 = bigger groups)
- **Analysis Type**: Power curve vs. single calculation
- **Significance Level**: Type I error rate (default: 0.05)

#### Simulation Parameters
- **Number of Simulations**: Balance between precision and speed (default: 1000)
- **Maximum Sample Size**: Budget/feasibility constraints
- **Effect Size**: Known biologically meaningful effect size
- **ICC Values**: Intra-class correlation for experimental hierarchy

### Step 5: Generate JSON Files

#### Enhanced Directory Selection
- **Browse Directory**: Interactive file browser with:
  - Quick shortcuts (Desktop, Documents, Downloads)
  - Directory navigation and folder creation
  - OS-specific file manager integration
- **Current Directory**: Use R's working directory
- **Custom Path**: Enter any valid directory path

#### Filename Customization
- **Optional Prefix**: Add custom prefix to all generated files
- **Real-time Preview**: See example filenames before download
- **Date Stamping**: Automatic date suffixes for version control

#### Dual File Saving
Files are saved to both:
1. Browser's download folder (standard download)
2. Your specified custom directory

#### Download Options
- **Individual Files**: Download each JSON file separately
- **Complete Package**: Download all three files in a ZIP archive

## Using Generated JSON Files

Once generated, use the JSON files with RMeDPower2 functions:

```{r use_json_files, eval=FALSE}
library(RMeDPower2)

# Load configurations from JSON files
design <- readDesign("RMeDesign_config_2024-01-15.json")
model <- readProbabilityModel("ProbabilityModel_config_2024-01-15.json") 
power_params <- readPowerParams("PowerParams_config_2024-01-15.json")

# Run power analysis
result <- calculatePower(data = your_data, 
                        design = design,
                        model = model, 
                        power_param = power_params)
```

## App Features and Benefits

### Data-Driven Configuration
- Column names automatically populated from uploaded data
- Data type detection for appropriate parameter suggestions
- Sample value preview for validation

### Interactive Validation
- Real-time parameter checking
- Built-in help text and examples
- Visual feedback for required vs. optional parameters

### Smart Defaults
- Sensible default values for all parameters
- Minimal configuration required for basic analyses
- Progressive disclosure of advanced options

### Enhanced User Experience
- Step-by-step guidance through complex configuration
- Distribution selection guidelines with examples
- Filename organization with custom prefixes

### Reproducible Workflows
- JSON files provide complete analysis documentation
- Version control through date stamping
- Easy sharing and collaboration

## Common Shiny App Workflows

### Workflow 1: Quick Start with Defaults
1. Upload data
2. Select response and condition columns
3. Accept default settings
4. Generate JSON files

This creates valid configurations using smart defaults, perfect for initial exploration.

### Workflow 2: Detailed Configuration
1. Upload data and review structure
2. Configure all RMeDesign parameters based on experimental structure
3. Select appropriate distribution based on data type
4. Customize power analysis parameters for specific goals
5. Set up organized file output with custom prefixes

### Workflow 3: Multiple Analyses
1. Configure base design and model
2. Create multiple PowerParams configurations for different scenarios:
   - Different target columns
   - Various effect sizes
   - Different simulation parameters
3. Use filename prefixes to organize related analyses

## Troubleshooting the Shiny App

### Common Issues
- **App won't start**: Ensure all packages installed with `source("install_packages.R")`
- **Data won't load**: Check file format (CSV/RDS) and permissions
- **Distribution dropdown hidden**: Fixed in v1.1 - dropdown appears when "Non-normal" selected
- **JSON validation errors**: Verify all required fields completed

### Best Practices
- Start with smaller datasets for faster loading
- Use data preview to verify column structure before configuration
- Test configurations with example data before large datasets
- Keep JSON files organized by project/experiment
- Use descriptive filename prefixes for easy identification

### Getting Help
- Built-in help text provides guidance for each parameter
- Distribution selection guide with biological examples
- Example configurations generated automatically
- Comprehensive error messages and notifications

## Advanced Features

### File Organization
The app supports systematic file organization:
- Custom directory selection with browsing
- Filename prefixes for project organization
- Date stamping for version control
- Batch download of all configurations

### Integration with Package Workflow
The Shiny app seamlessly integrates with the standard RMeDPower2 workflow:

```{r shiny_integration_workflow, eval=FALSE}
# 1. Create configurations with Shiny app
# 2. Load and verify configurations
design <- readDesign("my_study_RMeDesign_config_2024-01-15.json")
model <- readProbabilityModel("my_study_ProbabilityModel_config_2024-01-15.json")
power_params <- readPowerParams("my_study_PowerParams_config_2024-01-15.json")

# 3. Validate with diagnostic functions
diagnose_result <- diagnoseDataModel(data = your_data, design = design, model = model)

# 4. Run power analysis
power_result <- calculatePower(data = your_data, design = design, 
                               model = model, power_param = power_params)

# 5. Interpret and iterate as needed
```

This interactive approach makes RMeDPower2 accessible to users of all experience levels while maintaining the flexibility and power of the underlying statistical framework.