---
title: "RMeDPower2 Class Configuration Guide"
author: "RMeDPower2 Package"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning = FALSE, message = FALSE)
```

# Overview

RMeDPower2 uses three S4 class objects to structure experimental designs and power analysis parameters. This guide provides decision criteria for configuring each class based on your experimental setup.

The three core classes are:

- **RMeDesign**: Defines your experimental structure and statistical model specifications
- **ProbabilityModel**: Specifies the statistical distribution of your response variable  
- **PowerParams**: Configures power analysis parameters

---

# RMeDesign Class

The `RMeDesign` class defines your experimental structure and statistical model specifications.

## Required Parameters

### `response_column`
**What it is:** Name of your outcome/dependent variable column  
**How to choose:** This should be the quantitative measure you're analyzing

```{r response_column_examples}
# Examples:
response_column = "cell_size"        # Cell morphology study
response_column = "expression_level" # Gene expression analysis
response_column = "reaction_time"    # Behavioral study
```

### `condition_column` 
**What it is:** Your main predictor/treatment variable  
**How to choose:** The primary experimental factor you want to test

```{r condition_column_examples}
# Examples:
condition_column = "treatment"       # Drug vs control
condition_column = "genotype"       # Wild-type vs mutant
condition_column = "time_point"     # Longitudinal measurements
```

### `condition_is_categorical`
**Decision criteria:**

- `TRUE`: Treatment groups, genotypes, categorical time points
- `FALSE`: Continuous measurements like dose, age, time as numeric

### `experimental_columns`
**What it is:** Hierarchical grouping variables (most general to most specific)  
**How to order:** Always list from highest to lowest level

```{r experimental_columns_examples}
# Examples:
experimental_columns = c("experiment", "plate", "well")     # Lab hierarchy
experimental_columns = c("batch", "animal", "sample")       # Animal study
experimental_columns = c("study_site", "participant")       # Clinical trial
```

## Optional Parameters

### `covariate`
**When to include:** You have a confounding variable to control for

```{r covariate_examples}
covariate = "age"           # Control for age effects
covariate = "batch_date"    # Control for batch effects
covariate = "baseline"      # Control for pre-treatment values
```

### `covariate_is_categorical`
**Decision criteria:**

- `TRUE`: Gender, batch labels, categorical age groups
- `FALSE`: Continuous age, baseline measurements, numeric scores

### `crossed_columns`
**When to use:** Some experimental units appear in multiple higher-level groups  
**Example:** Same cell line used across multiple experiments

```{r crossed_columns_example}
experimental_columns = c("experiment", "cell_line")
crossed_columns = "cell_line"  # Cell lines repeat across experiments
```

### `include_interaction`
**When to use:** You expect the covariate effect varies by condition

**Examples:**
- Drug effect varies by age
- Treatment response differs by genotype  
- Time effects depend on intervention group

### `random_slope_variable`
**When to use:** You expect the effect varies across experimental units

**Options:**
- `"condition_column"`: Treatment effects vary across subjects/units
- `"covariate"`: Covariate effects vary across experimental levels

---

# ProbabilityModel Class

Defines the statistical distribution of your response variable.

## `error_is_non_normal`
**Decision criteria:**

- `FALSE`: Continuous, approximately normal data (heights, measurements, expression levels)
- `TRUE`: Count data, proportions, or clearly non-normal distributions

## `family_p` (when `error_is_non_normal = TRUE`)

### `"poisson"`
- **Use for:** Count data (number of cells, events, mutations)
- **Characteristics:** Non-negative integers, variance ≈ mean
- **Example:** Number of colonies per plate

### `"binomial"` 
- **Use for:** Proportions or binary outcomes
- **Requires:** `total_column` specifying denominator
- **Example:** Number of positive cells out of total counted

### `"negative_binomial"`
- **Use for:** Overdispersed count data 
- **When:** Count data where variance > mean
- **Example:** RNA-seq read counts

### `"Gamma"`
- **Use for:** Positive continuous data with right skew
- **Example:** Reaction times, concentrations

---

# PowerParams Class

Configures power analysis parameters.

## Core Parameters

### `target_columns`
**What it is:** Which experimental factors to vary in power analysis  
**How to choose:** The bottleneck factor(s) in your design

```{r target_columns_examples}
target_columns = "experiment"     # Want more experimental replicates
target_columns = "animal"         # Need more animals per group
target_columns = c("experiment", "animal")  # Optimize both levels
```

### `levels`
**Decision criteria for each target column:**

- `1`: Increase number of groups/levels (more experiments, more animals)
- `0`: Increase sample size within existing groups (more samples per animal)

### `power_curve`
- `1`: Generate power curves showing power vs. sample size
- `0`: Calculate power for specific sample size

## Sample Size Parameters

### `max_size`
**Default:** Current size × 5  
**When to specify:** You have budget/feasibility constraints

```{r max_size_example}
max_size = c(10, 50)  # Max 10 experiments, 50 animals per group
```

### `nsimn`
**Default:** 1000 simulations  
**When to change:**

- Increase for more precision (slower)
- Decrease for faster exploratory analysis

## Effect Size Specification

### `effect_size`
**When to specify:** You know the biologically meaningful effect size  
**How to determine:**

- From pilot studies
- From literature
- From biological significance criteria

### `ICC` (Intra-Class Correlation)
**When to use:** You have prior knowledge of experimental variability  
**How to specify:** Vector matching `experimental_columns` order

---

# Complete Configuration Examples

## Example 1: Cell Culture Experiment

```{r cell_culture_example}
library(RMeDPower2)

# Design: Test drug effect on cell size across experiments and plates
design <- new("RMeDesign",
  response_column = "cell_area",
  condition_column = "treatment",           # "control" vs "drug"
  condition_is_categorical = TRUE,
  experimental_columns = c("experiment", "plate"),
  crossed_columns = NULL,                   # Plates are nested in experiments
  covariate = "passage_number",
  covariate_is_categorical = FALSE,         # Continuous variable
  include_interaction = FALSE
)

model <- new("ProbabilityModel",
  error_is_non_normal = FALSE              # Cell area is approximately normal
)

power_params <- new("PowerParams",
  target_columns = "experiment",           # Want to know how many experiments needed
  levels = 1,                             # Add more experiments (not more plates per experiment)
  power_curve = 1,                        # Generate power curve
  nsimn = 1000,
  alpha = 0.05
)

# Run power analysis
# result <- calculatePower(data = your_data, design = design, 
#                          model = model, power_param = power_params)
```

## Example 2: RNA-seq Count Data

```{r rnaseq_example}
# Design: Compare gene expression between genotypes with batch effects
design <- new("RMeDesign",
  response_column = "read_count",
  condition_column = "genotype",
  condition_is_categorical = TRUE,
  experimental_columns = c("batch", "animal"),
  total_column = "total_reads",            # For normalization
  covariate = NULL
)

model <- new("ProbabilityModel",
  error_is_non_normal = TRUE,
  family_p = "negative_binomial"           # Overdispersed count data
)

power_params <- new("PowerParams",
  target_columns = "animal",
  levels = 1,                             # Add more animals
  power_curve = 1,
  alpha = 0.05
)

# Run power analysis
# result <- calculatePower(data = your_data, design = design, 
#                          model = model, power_param = power_params)
```

## Example 3: Behavioral Study with Continuous Predictor

```{r behavioral_example}
# Design: Test dose-response relationship in reaction time study
design <- new("RMeDesign",
  response_column = "reaction_time",
  condition_column = "dose",               # Continuous dose levels
  condition_is_categorical = FALSE,
  experimental_columns = c("session", "subject"),
  covariate = "age",
  covariate_is_categorical = FALSE,
  include_interaction = TRUE,              # Dose effect may vary by age
  random_slope_variable = "condition_column" # Dose effects vary by subject
)

model <- new("ProbabilityModel",
  error_is_non_normal = TRUE,
  family_p = "Gamma"                      # Reaction times are right-skewed
)

power_params <- new("PowerParams",
  target_columns = "subject",
  levels = 1,
  power_curve = 1,
  effect_size = 0.5                       # Known meaningful effect size
)

# Run power analysis
# result <- calculatePower(data = your_data, design = design, 
#                          model = model, power_param = power_params)
```

## Example 4: Clinical Trial with Categorical Covariate

```{r clinical_example}
# Design: Multi-site clinical trial with gender as covariate
design <- new("RMeDesign",
  response_column = "symptom_score",
  condition_column = "treatment_group",
  condition_is_categorical = TRUE,
  experimental_columns = c("site", "participant"),
  covariate = "gender",
  covariate_is_categorical = TRUE,         # Categorical covariate
  include_interaction = TRUE,              # Treatment effect may differ by gender
  crossed_columns = NULL
)

model <- new("ProbabilityModel",
  error_is_non_normal = FALSE              # Symptom scores approximately normal
)

power_params <- new("PowerParams",
  target_columns = c("site", "participant"),
  levels = c(1, 1),                       # Optimize both sites and participants
  power_curve = 1,
  nsimn = 500,                            # Reduce for faster computation
  alpha = 0.05
)

# Run power analysis
# result <- calculatePower(data = your_data, design = design, 
#                          model = model, power_param = power_params)
```

---

# Decision Flowchart

## Step 1: Data Structure
1. **What's your outcome variable?** → `response_column`
2. **What's your main comparison?** → `condition_column`
3. **What are your grouping levels?** → `experimental_columns`

## Step 2: Statistical Considerations
1. **Is your outcome normally distributed?** → `error_is_non_normal`
2. **Do you have covariates to control for?** → `covariate`, `include_interaction`
3. **Do effects vary across groups?** → `random_slope_variable`
4. **Is your covariate categorical or continuous?** → `covariate_is_categorical`

## Step 3: Power Analysis Goals
1. **What's your limiting factor?** → `target_columns`
2. **Do you want more groups or bigger groups?** → `levels`
3. **What's your budget constraint?** → `max_size`

---

# Tips for Success

## Data Preparation
- Ensure column names match exactly what you specify in the design objects
- Check that categorical variables are properly coded
- Verify hierarchical structure is correctly represented

## Model Selection
- Use diagnostic plots to verify distributional assumptions
- Consider the `diagnoseDataModel()` function to check model appropriateness
- Start with simpler models before adding interactions

## Power Analysis Strategy
- Begin with smaller `nsimn` values for exploratory analysis
- Focus on one `target_column` at a time initially
- Use pilot data or literature to inform effect size estimates

## Troubleshooting
- If power analysis fails, check that your design matches your data structure
- Ensure sufficient data for the complexity of your model
- Verify that crossed vs. nested experimental factors are correctly specified

---

# Additional Resources

## Package Functions
- `calculatePower()`: Main function for power analysis
- `diagnoseDataModel()`: Check model assumptions and identify outliers
- `get_model_and_data()`: Fit models and examine results
- `check_normality()`: Assess normality of residuals

## Input Templates
The package includes JSON templates in the `input_templates/` directory:
- `design_template.json`: Template for RMeDesign parameters
- `power_param_template.json`: Template for PowerParams parameters  
- `stat_model_template.json`: Template for ProbabilityModel parameters

## Example Data
Use the built-in example datasets (`RMeDPower_data1` through `RMeDPower_data7`) to practice with different experimental designs and learn the package functionality.

```{r load_example_data, eval=TRUE}
# Load and examine example data
library(RMeDPower2)
data(RMeDPower_data1)
head(RMeDPower_data1)
str(RMeDPower_data1)
```

This framework ensures your class objects accurately represent your experimental design and analysis goals, leading to meaningful and reliable power analysis results.