---
title: "Getting Started with RMeDPower2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with RMeDPower2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Introduction

RMeDPower2 is an R package for statistical power analysis in repeated measures experiments. It provides complete functionality to analyze data from hierarchical or crossed experimental designs, test modeling assumptions, identify outliers, estimate statistical power, and perform sample size calculations.

## Installation

You can install RMeDPower2 from GitHub:

```{r eval=FALSE}
# Install devtools if you haven't already
install.packages("devtools")

# Install RMeDPower2 from GitHub
devtools::install_github("gladstone-institutes/RMeDPower2", build_vignettes = TRUE)
```

## Quick Start

Here's a basic example of using RMeDPower2:

```{r eval=FALSE}
library(RMeDPower2)

# Load example data
data(RMeDPower_data1)

# Define experimental design
design <- new("RMeDesign",
    response_column = "cell_size2",
    condition_column = "classification",
    experimental_columns = c("experiment", "line"),
    condition_is_categorical = TRUE
)

# Define probability model
model <- new("ProbabilityModel",
    error_is_non_normal = FALSE
)

# Define power parameters
power_param <- new("PowerParams",
    target_columns = c("experiment"),
    power_curve = 1,
    nsimn = 100
)

# Step 1: Diagnose data and model assumptions
diagnose_res <- diagnoseDataModel(RMeDPower_data1, design, model)

# Step 2: Calculate power
power_res <- calculatePower(RMeDPower_data1, design, model, power_param)

# Step 3: Get parameter estimates
estimate_res <- getEstimatesOfInterest(RMeDPower_data1, design, model)
```

## Key Features

- **Experimental Design Support**: Handles nested and crossed experimental designs
- **Power Analysis**: Simulation-based power calculations using mixed-effects modeling
- **Outlier Detection**: Two-level outlier identification at observation and group levels
- **Model Validation**: Comprehensive assumption testing and diagnostic visualizations
- **Flexible Distributions**: Support for normal, binomial, Poisson, and negative binomial distributions

## Next Steps

- Read the [Tutorial](tutorial.html) for a complete walkthrough
- See [Examples](examples.html) for real-world applications
- Check the [Reference](../reference/) for detailed function documentation